---
title: "summative1_part2"
format: html
editor: visual
---

## Summative 1: Part 2

Your brief is to answer the following question: "Is it feasible to identify AI-averse individuals, without directly asking about their attitudes to this technology?"

Your answer should be a report of around 500-700 words that a) describes your approach to designing the model, b) demonstrates your final trained model's performance, and c) answers the brief providing a clear justification. You should think carefully about how you structure this report to present your findings in a clear manner. You may use (well-formatted) charts to help illustrate your claims. This report should be submitted as a .html file.


- Data Cleaning and Preprocessing: taking care of NA values
- Feature Selection and Engineering
- Normalisation and Scaling
- Exploratory Data Analysis: graphs/visuals/ correlations
- k mean clustering? Different categories of people
- Decision Trees: CART
- Evaluation



```{r}
library(haven)
# Load the data set
oxis_data <- haven::read_dta("UKDA-9146-stata/stata/stata14/oxis2019ukda.dta")
```


# Defining variables in each group
```{r}
common_vars <- c("age", "gender", "eth4", "marstat", 
                 "impinftv", "impinfrad", "impinfine", "impinfnews", "impenttv", 
                 "impentrad", "impentppl", "impentine", "relpnews", "reltv", "relrad", 
                 "relonews", "relsclm", "relsrch", "trmost", "funlit", "intnet", 
                 "agttry", "agtbet", "aganon", "agcred", "agpriv", "agtdate", "agpdata", 
                 "agai", "actsport", "actloc", "actunion", "actenv", "actpol", "actchar", 
                 "actrel", "p_direc", "p_pwill", "p_peop", "p_prun", "p_ptalk", "p_psell", 
                 "agtwout", "agtfail", "agtbre", "agtfun", "aglerdis", "aglerwant", "agoff", 
                 "aglereasy", "mob", "mobs", "mobmail2", "mobppix2", "mobspix2", "moblmus2", 
                 "mobdir2", "mobsns2", "mobtpix2", "mobtrav2", "mobform2", "mobttime", 
                 "mobvid", "mobstime", "mobfrust", "mobavoid", "mobwast", "mobptime", 
                 "mobeasy", "moblone", "mobesc", "mobtouch", "mobnews", "mobgamep", 
                 "moborg", "hhcabsat", "hhtab", "hhgame", "hhtv", "hhwear", "hhstrem", 
                 "hhpva", "ncomp", "usecomp", "usenet", 
                 "ageasy", "agfrust", "agimmor", "agaddict", "agexh", 
                 "agintouch", "agmeet", "agstime", "agpers", "agremv", "agwtime", "agptime", 
                 "aglonely", "agttime", "agenjoy", "agescape", "agpolite", "agvulg", 
                 "p_int", "p_party", "p_conmp", "p_pet", "p_buypol", "p_opinf", "p_cmt", 
                 "p_censor", "p_caresay", "p_hideinf", "p_freesay", "p_efqual", "p_efund", 
                 "p_efeasy", "p_efgood", "p_power", "p_moresay", "p_kwgov", "p_pubkw", 
                 "p_clim", "p_immig", 
                 "dgorig", "edntt", "labfor", "workstudy", "studywork", "workhome", 
                 "workuse", "workperf", "wad", "wcoll", "wdoc", "wmat", "wdobet", "wgetbet", 
                 "yrborn", "marstat", "adulthh", "kidhh", "disab", "disabnet", 
                 "urbrur", "income", "lackcomp", "leftout", "isolate", "agshy", "agoutgo", 
                 "agnocon", "agsucc", "agown", "agright", "agnoler", "agopp", "agideal", 
                 "agcare", "agdislife", "aginjus", "agwthink", "agwpay", "frgamb", "frsex")


## Internet Users
current_users_vars <- c("u_acchome", "u_accmob", "u_accsuw", "u_acclib", "u_accfre", 
                         "usedfir", "u_ability", "hhfast", "cs_bank", "cs_form", "cs_vid", 
                         "cs_pjt", "cs_heal", "u_snsfb", "u_snsli", "u_snstw", "u_snsdat", 
                         "u_snspin", "u_snsgoo", "u_snsins", "u_snssna", "u_snsvid", 
                         "u_snsoth", "u_frstat", "u_frspix", "u_frscmt", "u_frrwri", 
                         "u_frnfo", "u_frcont", "u_frpcmt", "u_frsper3", "u_frswri3", 
                         "u_frpriv3", "u_frpol3", "u_frscl3", "u_frslikecom3", "u_frdrop3", 
                         "u_frups3", "u_frrpcmt3", "u_agwast", "u_agptime", "bl_offy", 
                         "bl_arg", "u_heasy", "u_gofir", "u_hfam", "u_hsuw", "u_hlib", 
                         "u_hvorg", "u_htrain", "u_hvid", "u_hother", "u_frattach", 
                         "u_frim", "u_frsns", "u_frcalls", "u_frrblog", "u_frwblog", 
                         "u_frsite", "u_frcompl", "u_frfloc", "u_frpprob", "u_time", 
                         "u_frppix", "u_frrpix", "u_frwri", "u_frpvid", "u_frrvid", 
                         "u_cmtpr2", "u_cmtser2", "u_cmtusr2", "s_freq", "s_wheng", 
                         "s_mpt", "s_cg", "s_wro", "s_lern", "s_over", "s_nthk", "s_chek", 
                         "pn_fam", "pn_pol", "pn_tv", "pn_rad", "pn_fnews", "pn_onews", 
                         "pn_sns", "pn_srch", "pn_disa", "pn_dif", "pn_conf", "pn_med", 
                         "u_frogame", "u_frngame", "u_frdlmus", "u_frlmus", "u_frdlvid", 
                         "u_frwmov", "u_frlpix", "u_frwtv", "u_frpod", "u_frbook", 
                         "u_comaddr", "u_comdob", "u_compho", "u_comcred", "u_compix", 
                         "u_fund", "u_frnews", "u_frevent", "u_frsport", "u_frtrav", 
                         "u_frgoog", "u_frpopnews", "u_ijob", "u_smon", "u_fevent", 
                         "u_fheal", "u_bmed", "u_metper", "u_pubann", "u_fjob", "u_conrely", 
                         "u_convirus", "u_conpix", "u_conmus", "u_padd2", 
                         "u_pshop2", "u_pmed2", "u_pmar2", "u_page2", "u_pprot2","u_hnone", "u_confren",
                         "u_frfact", "u_frproj", "u_frint", "u_frdiy", "u_lertran", 
                         "u_lertfree", "u_lertbuy", "u_lertobuy", "u_lerloc", 
                         "u_lerliv", "u_leroff", "u_leronl", "u_lernneed", "u_lerncost", 
                         "u_lerntime", "u_lernwher", "u_lernwant", "u_lernwhat", "u_lernint", 
                         "u_lerprod", "u_lersskil", "u_lercmm", "u_frbuy", "u_frbills", 
                         "u_frbank", "u_frcomp", "u_frfood", "u_frsell", "locserv", 
                         "loctax", "govserv", "govtax", "sch", "mp", "policy", "govweb", 
                         "u_conad", "u_convir", "u_concom", "u_conpay", "u_actad", 
                         "u_actvirus", "u_actcom", "u_spam", "u_virus", "u_misrep", 
                         "u_stolen", "u_hate", "u_theft", "u_hack", "u_actpay", "u_actspw", 
                         "u_actcpw", "u_actsec", "frsport", "frtv", "frread", "frgoout", 
                         "frcards", "frclass")



## Never Used Internet
never_used_vars <- c("n_reint", "n_reacc", "n_recomp", "n_redif", "n_reuseful", "n_reexpen", 
                     "n_repriv", "n_rebad", "n_renotime", "n_renoint", "n_reusenet", 
                     "n_retime", "n_reage", "n_relikeme", "n_remoney", "n_redetail", "n_reimp",
                     "n_agmiss", "n_hpar", "n_hfrien", "n_hpart", "n_hkid", "n_hsib", "n_hlib", 
                     "n_hcmy", "n_hcoll", "n_hpay", "n_agleftout", "n_agbetnot", "n_agfuture", 
                     "n_agperbet", "n_hsome", "n_par", "n_frien", "n_part", "n_kid", "n_sib", 
                     "n_lib", "n_cmy", "n_coll", "n_pay", "n_future", "n_gsave", "n_gfam", 
                     "n_gint", "n_gtv", "n_gbank", "n_gben", "n_gjob", "ynacc")

## Ex-Users
ex_user_vars <- c("e_retry", "e_rework", "e_resch", "e_rerec", "e_reacc", "e_rehkid", "e_reintouch",
                  "e_buysvc", "e_reimpu", "e_reint", "e_remove", "e_recomp", "e_redif", "e_reuseful", 
                  "e_reexpen", "e_repriv", "e_rebad", "e_renotime", "e_renoint", "e_rehow", 
                  "e_retime", "e_reage", "e_relikeme", "e_rejob", "e_reimps", "e_renotuse")


```


#Filtering Data to create Groups
```{r}
never_used <- oxis_data %>% 
  select(all_of(c(common_vars, never_used_vars)))

ex_users <- oxis_data %>% 
  select(all_of(c(common_vars, ex_user_vars)))

current_users <- oxis_data %>% 
  select(all_of(c(common_vars, current_users_vars)))

```


#Current Users: Fitting CART model

```{r}
# Load necessary libraries
library(haven)       # Reading STATA files
library(dplyr)       # Data manipulation
library(rpart)       # CART model
library(rpart.plot)  # Decision tree visualization
library(caret)     # Model evaluation
library(ROSE)        # Class balancing if needed
library(performanceEstimation)       # SMOTE for class balancing

# 🌟 Train & Prune Decision Tree for Internet Users with Cross-Validation

# Step 1: Prepare Data
current_users <- oxis_data %>%
  select(all_of(c(common_vars, current_users_vars))) %>%  
  drop_na(agai) %>%                                       
  filter(agai %in% c(1, 2, 4, 5)) %>%                    
  mutate(
    agai = factor(ifelse(agai %in% c(1, 2), 1, 0), levels = c(0, 1))
  ) %>%
  mutate(agai = relevel(agai, ref = "1"))  # Set AI-Averse (1) as positive class


# Train-Test Split (80% Training, 20% Testing)
train_indices <- sample(1:nrow(current_users), size = 0.8 * nrow(current_users))
train_data <- current_users[train_indices, ]
test_data  <- current_users[-train_indices, ]

# Step 3: Define Cost Matrix (Higher penalty for misclassifying AI-Averse (1))
loss_matrix <- matrix(c(0, 10,  # Cost of correctly classifying AI-Pro (0) = 0, Misclassifying AI-Averse (1) = 5
                        1, 0), # Cost of Misclassifying AI-Pro (0) = 1, Correctly classifying AI-Averse (1) = 0
                      byrow = TRUE, nrow = 2)


cart_model <- rpart(agai ~ . , 
                         data = train_data, 
                         method = "class", 
                         parms = list(loss= loss_matrix, split="information"), # Apply cost matrix
                         cp = 0.001, 
                         minsplit = 10, 
                         minbucket =3)

# Check Complexity Parameter (cp) Table
printcp(cart_model)

# Step 5: Prune the Tree
best_cp <- cart_model$cptable[which.min(cart_model$cptable[, "xerror"]), "CP"] * 0.5
pruned_cart <- prune(cart_model, cp = best_cp)

# Step 6: Visualize the Pruned Tree
rpart.plot(pruned_cart, type = 3, extra = 101, tweak = 1.2)

# Step 7: Model Evaluation on Test Data
y_pred <- predict(pruned_cart, newdata = test_data, type = "class")
conf_matrix <- confusionMatrix(y_pred, test_data$agai, positive = "1")

# Print Model Performance Metrics
print(conf_matrix)

library(caret)
var_imp <- varImp(cart_model)
print(var_imp)

# Separate AI-averse and AI-positive cases
ai_averse <- train_data %>% filter(agai == 1)
ai_positive <- train_data %>% filter(agai == 0)

# Oversample AI-averse cases by duplicating them
ai_averse_oversampled <- ai_averse[sample(1:nrow(ai_averse), size = nrow(ai_averse) * 2, replace = TRUE), ]

# Combine back into a new balanced training dataset
train_data_balanced <- bind_rows(ai_averse_oversampled, ai_positive)

# Check class distribution after oversampling
table(train_data_balanced$agai)

# Train Decision Tree on Balanced Data
cart_model_balanced <- rpart(agai ~ ., 
                             data = train_data_balanced, 
                             method = "class", 
                             cp = 0.001, 
                             minsplit = 10, 
                             minbucket = 3)

# Evaluate Model on Test Data
y_pred_balanced <- predict(cart_model_balanced, newdata = test_data, type = "class")
conf_matrix_balanced <- confusionMatrix(y_pred_balanced, test_data$agai, positive = "1")

# Print Performance Metrics
print(conf_matrix_balanced)




```


```{r}
cluster_data <- current_users %>%
  select(age, intnet, agshy, agtbet, agpriv, agwthink, moborg, cs_bank) %>%
  na.omit()


library(factoextra)

fviz_nbclust(cluster_data, kmeans, method = "wss")  # Elbow method


set.seed(42)
kmeans_result <- kmeans(cluster_data, centers = 3, nstart = 25)

# Add cluster labels to data
cluster_data$cluster <- factor(kmeans_result$cluster)

# Visualize Clusters (2D Projection)
ggplot(cluster_data, aes(x = age, y = agtbet, color = cluster)) +
  geom_point(alpha = 0.6) +
  labs(title = "K-Means Clustering of AI-Averse Groups",
       x = "Age", y = "Belief AI Will Benefit Society") +
  theme_minimal()

```






```{r}

#Tree for internet users

# Load necessary libraries
library(haven)       # For reading STATA files
library(dplyr)       # For data manipulation
library(rpart)       # CART model
library(rpart.plot)  # For visualizing the tree
library(caret)       # For model evaluation

# 🌟 Train CART Model for Internet Users
cart_model_current <- oxis_data %>%
  select(all_of(c(common_vars, current_users_vars))) %>%   # Select relevant variables
  drop_na(agai) %>%                                       # Remove NAs in target variable
  filter(agai %in% c(1, 2, 4, 5)) %>%                     # Remove "Don't Know" (-3) & Neutral (3)
  mutate(
    agai = factor(ifelse(agai %in% c(1, 2), 1, 0),        # Convert agai to binary: 1 = AI-Averse, 0 = AI-Pro
                  levels = c(0, 1))                       # Ensure correct factor levels
  ) %>%
  mutate(agai = relevel(agai, ref = "1")) %>%             # Set AI-Averse (1) as the positive class
  rpart(agai ~ ., data = ., method = "class")             # Train the CART model

# 🌟 Plot the Decision Tree
rpart.plot(cart_model_current, type = 3, extra = 101, tweak = 1.2)

# 🌟 Model Evaluation - Confusion Matrix
y_pred_current <- predict(cart_model_current, newdata = oxis_data, type = "class")
conf_matrix_current <- confusionMatrix(y_pred_current, factor(oxis_data$agai, levels = c(0, 1)), positive = "1")

# Print Results
print(conf_matrix_current)
printcp(cart_model_current)  # Check tree complexity
summary(cart_model_current)   # Print tree summary

```



#Ex Users: Tree-Based Model

```{r}

#Ex Users

# Load necessary libraries
library(haven)       # For reading STATA files
library(dplyr)       # For data manipulation
library(rpart)       # CART model
library(rpart.plot)  # For visualizing the tree
library(caret)       # For model evaluation
library(tidyverse)



# Step 1: Prepare Data
ex_users <- oxis_data %>%
  select(all_of(c(common_vars, ex_user_vars))) %>%  
  drop_na(agai) %>%                                       
  filter(agai %in% c(1, 2, 4, 5)) %>%                    
  mutate(
    agai = factor(ifelse(agai %in% c(1, 2), 1, 0), levels = c(0, 1))
  ) %>%
  mutate(agai = relevel(agai, ref = "1"))  # Set AI-Averse (1) as positive class



# Train-Test Split (80% Training, 20% Testing)
train_indices <- sample(1:nrow(ex_users), size = 0.8 * nrow(ex_users))
train_data <- ex_users[train_indices, ]
test_data  <- ex_users[-train_indices, ]

# Step 3: Define Cost Matrix (Higher penalty for misclassifying AI-Averse (1))
loss_matrix <- matrix(c(0, 5,  # Cost of correctly classifying AI-Pro (0) = 0, Misclassifying AI-Averse (1) = 5
                        1, 0), # Cost of Misclassifying AI-Pro (0) = 1, Correctly classifying AI-Averse (1) = 0
                      byrow = TRUE, nrow = 2)


cart_model <- rpart(agai ~ . , 
                         data = train_data, 
                         method = "class", 
                         parms = list(split="information"), # Apply cost matrix
                         cp = 0.001, 
                         minsplit = 10, 
                         minbucket =3)

# Check Complexity Parameter (cp) Table
printcp(cart_model)

# Step 5: Prune the Tree
best_cp <- cart_model$cptable[which.min(cart_model$cptable[, "xerror"]), "CP"] * 0.5
pruned_cart <- prune(cart_model, cp = best_cp)

# Step 6: Visualize the Pruned Tree
rpart.plot(pruned_cart, type = 3, extra = 101, tweak = 1.2)

# Step 7: Model Evaluation on Test Data
y_pred <- predict(pruned_cart, newdata = test_data, type = "class")
conf_matrix <- confusionMatrix(y_pred, test_data$agai, positive = "1")

# Print Model Performance Metrics
print(conf_matrix)


```

# Non Users: Tree-Based Model

```{r}
#Non Users

# Step 1: Prepare Data
non_users <- oxis_data %>%
  select(all_of(c(common_vars, never_used_vars))) %>%  
  drop_na(agai) %>%                                       
  filter(agai %in% c(1, 2, 4, 5)) %>%                    
  mutate(
    agai = factor(ifelse(agai %in% c(1, 2), 1, 0), levels = c(0, 1))
  ) %>%
  mutate(agai = relevel(agai, ref = "1"))  # Set AI-Averse (1) as positive class



# Train-Test Split (80% Training, 20% Testing)
train_indices <- sample(1:nrow(non_users), size = 0.8 * nrow(non_users))
train_data <- non_users[train_indices, ]
test_data  <- non_users[-train_indices, ]

# Step 3: Define Cost Matrix (Higher penalty for misclassifying AI-Averse (1))
loss_matrix <- matrix(c(0, 5,  # Cost of correctly classifying AI-Pro (0) = 0, Misclassifying AI-Averse (1) = 5
                        1, 0), # Cost of Misclassifying AI-Pro (0) = 1, Correctly classifying AI-Averse (1) = 0
                      byrow = TRUE, nrow = 2)


cart_model <- rpart(agai ~ ., 
                         data = train_data, 
                         method = "class", 
                         parms = list(split="information"), # Apply cost matrix
                         cp = 0.001, 
                         minsplit = 10, 
                         minbucket =3)

# Check Complexity Parameter (cp) Table
printcp(cart_model)

# Step 5: Prune the Tree
best_cp <- cart_model$cptable[which.min(cart_model$cptable[, "xerror"]), "CP"] * 0.5
pruned_cart <- prune(cart_model, cp = best_cp)

# Step 6: Visualize the Pruned Tree
rpart.plot(pruned_cart, type = 3, extra = 101, tweak = 1.2)

# Step 7: Model Evaluation on Test Data
y_pred <- predict(pruned_cart, newdata = test_data, type = "class")
conf_matrix <- confusionMatrix(y_pred, test_data$agai, positive = "1")

# Print Model Performance Metrics
print(conf_matrix)

```

```{r}
#Non Users

# Step 1: Prepare Data
non_users <- oxis_data %>%
  select(all_of(c(common_vars, never_used_vars))) %>%  
  drop_na(agai) %>%                                       
  filter(agai %in% c(1, 2, 4, 5)) %>%                    
  mutate(
    agai = factor(ifelse(agai %in% c(1, 2), 1, 0), levels = c(0, 1))
  ) %>%
  mutate(agai = relevel(agai, ref = "1"))  # Set AI-Averse (1) as positive class



# Train-Test Split (80% Training, 20% Testing)
train_indices <- sample(1:nrow(non_users), size = 0.8 * nrow(non_users))
train_data <- non_users[train_indices, ]
test_data  <- non_users[-train_indices, ]

# Step 3: Define Cost Matrix (Higher penalty for misclassifying AI-Averse (1))
loss_matrix <- matrix(c(0, 5,  # Cost of correctly classifying AI-Pro (0) = 0, Misclassifying AI-Averse (1) = 5
                        1, 0), # Cost of Misclassifying AI-Pro (0) = 1, Correctly classifying AI-Averse (1) = 0
                      byrow = TRUE, nrow = 2)


cart_model <- rpart(agai ~ ., 
                         data = train_data, 
                         method = "class",  # Apply cost matrix
                         cp = 0.001, 
                         minsplit = 10, 
                         minbucket =3)

# Check Complexity Parameter (cp) Table
printcp(cart_model)

# Step 5: Prune the Tree
best_cp <- cart_model$cptable[which.min(cart_model$cptable[, "xerror"]), "CP"] * 0.5
pruned_cart <- prune(cart_model, cp = best_cp)

# Step 6: Visualize the Pruned Tree
rpart.plot(pruned_cart, type = 3, extra = 101, tweak = 1.2)

# Step 7: Model Evaluation on Test Data
y_pred <- predict(pruned_cart, newdata = test_data, type = "class")
conf_matrix <- confusionMatrix(y_pred, test_data$agai, positive = "1")

# Print Model Performance Metrics
print(conf_matrix)

```







